A non-contextual hidden variable model for QM

### 2016-07-06 finalPaper.tex
added another proof for the determinism theorem, fixed a few other things and updated the git ignore

\begin{notn} (a) $\psi\in\mathcal{H}$ would typically represent the quantum mechanical state of the system (assumed pure), residing in the Hilbert space $\mathcal{H}$, (b) $\hat{\mathcal{H}}$ is defined to mean $\mathcal{H}\otimes\mathcal{H}^{\dagger}$ (i.e. the set of Hermitian operators, observables, on the space $\mathcal{H}$), (c) $[\mathcal{H}]$ is defined to mean $(\mathcal{H},\,\mathbb{R}^{\otimes})$, which represents the state of the system including hidden variables, (d) $[\psi]\in[\mathcal{H}]$ will represent the state of the system, including hidden variables, (e) a prediction map is $m:\hat{\mathcal{H}},[\mathcal{H}]\to\mathbb{R}$, (f) a sequence map is $s:\hat{\mathcal{H}},[\mathcal{H}],\mathbb{R}\to[\mathcal{H}]$, (g) the set of all experimental setups is denoted by $\mathcal{E}$, (h) a setup map is $e:\hat{\mathcal{H}}\to\mathcal{E}$, (i) number of observables is represented by $N \in \mathbb{Z}^+$, (j) $f$ is an arbitrary map from $\hat{\mathcal{H}}^{\otimes N}\to \hat{\mathcal{H}}$ constructed using multiplication and addition of the observables, and multiplication with complex numbers, (k) $\tilde{f}$ is a map from $\mathbb{R}^{\otimes N} \to \mathbb{R}$ constructed by replacing observables in $f$ with real numbers.
\end{notn}

@@

Before proceeding we note, however, that QM enforces \emph{sequential
multiplicativity} at least in certain cases. This result will be used
to assess accuracy of any hidden variable model we might propose.
\begin{prop*}
Let the system be in a state, s.t. measurement of $\hat{C}$ yields
repeatable results (same result each time). Then according to QM,
\emph{sequential multiplicativity} holds, where $\hat{C}\equiv f(\hat{B}_{1},\hat{B}_{2},\dots\hat{B}_{n})$,
and $\hat{B}_{i}$ are as defined. \end{prop*}
\begin{proof}
Assume without loss of generality that $\hat{B}_{1},\hat{B}_{2},\dots\hat{B}_{n}$
are mutually compatible (commuting) and complete set of operators.
If say for instance the set is not complete, then one can add the
missing operators and label them as aforesaid. It follows that $\exists$
$\left|\mathbf{b}=\left(b_{1}^{(l_{1})},b_{2}^{(l_{2})},\dots b_{n}^{(l_{n})}\right)\right\rangle $
s.t. $\hat{B}_{i}\left|\mathbf{b}\right\rangle =b_{i}^{(l_{i})}\left|\mathbf{b}\right\rangle $,
where $l_{i}$ indexes the eigenvalues corresponding to $\hat{B}_{i}$
and that $\sum_{\mathbf{b}}\left|\mathbf{b}\right\rangle \left\langle \mathbf{b}\right|=\hat{\mathbb{I}}$.
Let the state of the system be given by $\left|\psi\right\rangle $
and it must be s.t. $\hat{C}\left|\psi\right\rangle =c\left|\psi\right\rangle $,
by assumption. For the statement to follow, one need only show that
$\left|\psi\right\rangle $ must be made of only those $\left|\mathbf{b}\right\rangle $s,
which satisfy $c=\tilde f(b_{1}^{(l_{1})},b_{2}^{(l_{2})},\dots b_{n}^{(l_{n})})$.
This is the crucial step. Proving this is so is trivial. We start
with $\hat{C}\left|\psi\right\rangle =c\left|\psi\right\rangle $
and take its inner product with $\left\langle \mathbf{b}\right|$
to get 
\begin{eqnarray*}
\left\langle \mathbf{b}\right|\hat{C}\left|\psi\right\rangle  & = & c\left\langle \mathbf{b}|\psi\right\rangle ,\\
\left\langle \mathbf{b}\right|f(\hat{B}_{1},\hat{B}_{2},\dots\hat{B}_{n})\left|\psi\right\rangle  & = & c\left\langle \mathbf{b}|\psi\right\rangle ,\\
\tilde f(b_{1}^{(l_{1})},b_{2}^{(l_{2})},\dots b_{n}^{(l_{n})})\left\langle \mathbf{b}|\psi\right\rangle  & = & c\left\langle \mathbf{b}|\psi\right\rangle .
\end{eqnarray*}
Also, we have $\left|\psi\right\rangle =\sum_{\mathbf{b}}\left\langle \mathbf{b}|\psi\right\rangle \left|\mathbf{b}\right\rangle $,
from completeness. If we consider $\left|\mathbf{b}\right\rangle $s
for which $\left\langle \mathbf{b}|\psi\right\rangle \neq0$, then
we can conclude that indeed $c=f(b_{1}^{(l_{1})},b_{2}^{(l_{2})},\dots b_{n}^{(l_{n})})$.
However, when $\left\langle \mathbf{b}|\psi\right\rangle =0$, viz.
$\left|\mathbf{b}\right\rangle $s that are orthogonal to $\left|\psi\right\rangle $,
then nothing can be said. We can thus conclude that $\left|\psi\right\rangle $
is made only of those $\left|\mathbf{b}\right\rangle $s that satisfy
the required relation. That completes the proof.
\end{proof}
It is worth noting that in the PM case, where $\hat{R}_{i}$ and $\hat{C}_{j}$
are just $\pm\hat{\mathbb{I}}$, it follows that all states are their
eigenstates. Consequently, for these operators \emph{sequential multiplicativity}
must always hold.

@@

TODO: Complete this part.
\begin{appendix}
\section{Proofs of the two theorems}
For \thmref{det} we have the following proofs.
\begin{proof}
Bell construction ($\left|\mathcal{H}\right|\ge4$): Consider an operator
$\hat{B}:=a_{1}\otimes b_{1}+a_{2}\otimes b_{1}+a_{1}\otimes b_{2}-a_{2}\otimes b_{2}$
s.t. $-1\le\hat{a}_{i},\hat{b}_{j}\le1$ $\forall\,i,j\in\{1,2,3,4\}$
(which means that the spectrum of $\hat{a}$ and $\hat{b}$ are bounded
above and below). If we assume that a map $m$ of the aforesaid type
exists, then it can be shown that $\left\langle \hat{B}\right\rangle \le2$.
To see this, consider $\hat{A}_{1}=\hat{a}_{1}\otimes\hat{\mathbb{I}}$
and $\hat{A}_{2}=\hat{\mathbb{I}}\otimes\hat{b}_{1}$, so that if
$\hat{C}:=\hat{A}_{1}\hat{A}_{2}$ then property (2) entails that
$m(\hat{C})=m(\hat{A}_{1})m(\hat{A}_{2})$, viz. $m(\hat{a}_{1}\otimes\hat{b}_{1})=m(\hat{a}_{1}\otimes\hat{\mathbb{I}})m(\hat{\mathbb{I}}\otimes\hat{b}_{1})$.
Note that $m(\hat{*})$, may depend on the state $\left|\psi\right\rangle $
and the hidden variables%
\begin{comment}
, where $\hat{*}$ is taken to represent $\hat{B}_{k}$s which can
be generalised to all $\hat{a}_{i}$s and $\hat{b}_{j}$s
\end{comment}
. Therefore, we average over multiple measurements, which corresponds
to averaging over the hidden variables. This is expressed by $\left\langle m(\hat{a}_{1}\otimes\hat{b}_{1})\right\rangle =\left\langle m(\hat{a}_{1}\otimes\hat{\mathbb{I}})\right\rangle \left\langle m(\hat{\mathbb{I}}\otimes\hat{b}_{1})\right\rangle $.
One can repeat this argument for each term in $\left\langle \hat{B}\right\rangle $.
Using the fact that $-1\le m(\hat{*})\le1$, and that the extrema
will occur only at the extreme values of $m$, one can plugin $\pm1$
for each $\left\langle m(\hat{*})\right\rangle $ to obtain $\left\langle \hat{B}\right\rangle \le2$.
Quantum mechanically, one can show that $\left\langle \hat{B}\right\rangle =2\sqrt{2}$
for the state $\left|\psi\right\rangle =\left(\left|+-\right\rangle -\left|-+\right\rangle \right)/\sqrt{2}$,
where $\hat{\sigma}_{x}\left|\pm\right\rangle =\pm\left|\pm\right\rangle $,
and $\hat{\sigma}_{x,y,z}$ are the Pauli matrices, given in the z-basis
as 
\[
\hat{\sigma}_{x}\doteq\left[\begin{array}{cc}
0 & 1\\
1 & 0
\end{array}\right],\,\hat{\sigma}_{y}\doteq\left[\begin{array}{cc}
0 & -i\\
i & 0
\end{array}\right],\,\hat{\sigma}_{z}\doteq\left[\begin{array}{cc}
1 & 0\\
0 & -1
\end{array}\right].
\]
The observables are $\hat{a}_{1}=\hat{\sigma}_{z}$ and $\hat{a}_{2}=\hat{\sigma}_{x}$,
while $\hat{b}_{1}=-\frac{\hat{\sigma}_{z}+\hat{\sigma}_{x}}{\sqrt{2}}$
and $\hat{b}_{2}=\frac{\hat{\sigma}_{z}-\hat{\sigma}_{x}}{\sqrt{2}}$.
Since quantum mechanics violates the inequality, $\left\langle \hat{B}\right\rangle \le2$,
we conclude that the map $m$ can not simultaneously exist and be
consistent with QM.
\end{proof}

>>>>

\begin{notn} (a) $\psi\in\mathcal{H}$ would typically represent the quantum mechanical state of the system (assumed pure), residing in the Hilbert space $\mathcal{H}$, (b) $\hat{\mathcal{H}}$ is defined to mean $\mathcal{H}\otimes\mathcal{H}^{\dagger}$ (i.e. the set of Hermitian operators, observables, on the space $\mathcal{H}$), (c) $[\mathcal{H}]$ is defined to mean $(\mathcal{H},\,\mathbb{R}^{\otimes})$, which represents the state of the system including hidden variables, (d) $[\psi]\in[\mathcal{H}]$ will represent the state of the system, including hidden variables, (e) a prediction map is $m:\hat{\mathcal{H}},[\mathcal{H}]\to\mathbb{R}$, (f) a sequence map is $s:\hat{\mathcal{H}},[\mathcal{H}],\mathbb{R}\to[\mathcal{H}]$, (g) the set of all experimental setups is denoted by $\mathcal{E}$, (h) a setup map is $e:\hat{\mathcal{H}}\to\mathcal{E}$, (i) number of observables is represented by $N \in \mathbb{Z}^+$, (j) $f$ is an arbitrary map from $\hat{\mathcal{H}}^{\otimes N}\to \hat{\mathcal{H}}$ constructed using multiplication and addition of the observables, and multiplication with complex numbers\footnote{Strictly, the map $f$ would depend on the operators to ensure preservation of Hermiticity; thus a given $f$ will be defined only for a subset of $\hat {\mathcal H}^{\otimes N}$.}, (k) $\tilde{f}$ is a map from $\mathbb{R}^{\otimes N} \to \mathbb{R}$ constructed by replacing observables in $f$ with real numbers\footnote{$\tilde f$ constructed this way should strictly be written as $\tilde f:\mathbb{R}^{\otimes N} \to \mathbb{C}$.}.
\end{notn}

@@

Before proceeding we note, however, that QM enforces \emph{sequential
multiplicativity} at least in certain cases. This result will be used
to assess accuracy of any hidden variable model we might propose.
\begin{prop*}
Let the system be in a state, s.t. measurement of $\hat{C}$ yields
repeatable results (same result each time). Then according to QM,
\emph{sequential multiplicativity} holds, where $\hat{C}\equiv f(\hat{B}_{1},\hat{B}_{2},\dots\hat{B}_{n})$,
and $\hat{B}_{i}$ are as defined. \end{prop*}
\begin{proof}
Assume without loss of generality that $\hat{B}_{1},\hat{B}_{2},\dots\hat{B}_{n}$
are mutually compatible (commuting) and complete set of operators.
If say for instance the set is not complete, then one can add the
missing operators and label them as aforesaid. It follows that $\exists$
$\left|\mathbf{b}=\left(b_{1}^{(l_{1})},b_{2}^{(l_{2})},\dots b_{n}^{(l_{n})}\right)\right\rangle $
s.t. $\hat{B}_{i}\left|\mathbf{b}\right\rangle =b_{i}^{(l_{i})}\left|\mathbf{b}\right\rangle $,
where $l_{i}$ indexes the eigenvalues corresponding to $\hat{B}_{i}$
and that $\sum_{\mathbf{b}}\left|\mathbf{b}\right\rangle \left\langle \mathbf{b}\right|=\hat{\mathbb{I}}$.
Let the state of the system be given by $\left|\psi\right\rangle $
and it must be s.t. $\hat{C}\left|\psi\right\rangle =c\left|\psi\right\rangle $,
by assumption. For the statement to follow, one need only show that
$\left|\psi\right\rangle $ must be made of only those $\left|\mathbf{b}\right\rangle $s,
which satisfy $c=\tilde f(b_{1}^{(l_{1})},b_{2}^{(l_{2})},\dots b_{n}^{(l_{n})})$.
This is the crucial step. Proving this is so is trivial. We start
with $\hat{C}\left|\psi\right\rangle =c\left|\psi\right\rangle $
and take its inner product with $\left\langle \mathbf{b}\right|$
to get 
\begin{eqnarray*}
\left\langle \mathbf{b}\right|\hat{C}\left|\psi\right\rangle  & = & c\left\langle \mathbf{b}|\psi\right\rangle ,\\
\left\langle \mathbf{b}\right|f(\hat{B}_{1},\hat{B}_{2},\dots\hat{B}_{n})\left|\psi\right\rangle  & = & c\left\langle \mathbf{b}|\psi\right\rangle ,\\
\tilde f(b_{1}^{(l_{1})},b_{2}^{(l_{2})},\dots b_{n}^{(l_{n})})\left\langle \mathbf{b}|\psi\right\rangle  & = & c\left\langle \mathbf{b}|\psi\right\rangle .
\end{eqnarray*}
Also, we have $\left|\psi\right\rangle =\sum_{\mathbf{b}}\left\langle \mathbf{b}|\psi\right\rangle \left|\mathbf{b}\right\rangle $,
from completeness. If we consider $\left|\mathbf{b}\right\rangle $s
for which $\left\langle \mathbf{b}|\psi\right\rangle \neq0$, then we can conclude that indeed $c=\tilde f(b_{1}^{(l_{1})},b_{2}^{(l_{2})},\dots b_{n}^{(l_{n})})$.
However, when $\left\langle \mathbf{b}|\psi\right\rangle =0$, viz.
$\left|\mathbf{b}\right\rangle $s that are orthogonal to $\left|\psi\right\rangle $,
then nothing can be said. We can thus conclude that $\left|\psi\right\rangle $
is made only of those $\left|\mathbf{b}\right\rangle $s that satisfy
the required relation. That completes the proof.
\end{proof}
It is worth noting that in the PM case, where $\hat{R}_{i}$ and $\hat{C}_{j}$
are just $\pm\hat{\mathbb{I}}$, it follows that all states are their
eigenstates. Consequently, for these operators \emph{sequential multiplicativity}
must always hold.

@@

TODO: Complete this part.
\begin{appendix}
\section{Proofs of the two theorems}
For \thmref{det} we have the following proofs.
\begin{proof}
Trivial yet most general proof ($\left|\mathcal{H}\right|\ge2$): Consider $\hat A _1 = \sigma_y$, $\hat A_2 =\sigma_x$, $\hat C = f(\hat A_1,\hat A_2) = i\hat A_1 \hat A_2=\sigma_z$. Now we know that $m(\hat \sigma_j)=\pm1$ which entails that $m(\hat C)=  im(\hat A_1)m(\hat A_2)=i$ where we have used property (b). However, $m(\hat C)=m(\hat \sigma_z)=\pm1$ which is a contradiction. Thus no consistent $m$ can exist.
\end{proof}

\begin{proof}
Bell construction ($\left|\mathcal{H}\right|\ge4$): Consider an operator
$\hat{B}:=a_{1}\otimes b_{1}+a_{2}\otimes b_{1}+a_{1}\otimes b_{2}-a_{2}\otimes b_{2}$
s.t. $-1\le\hat{a}_{i},\hat{b}_{j}\le1$ $\forall\,i,j\in\{1,2,3,4\}$
(which means that the spectrum of $\hat{a}$ and $\hat{b}$ are bounded
above and below). If we assume that a map $m$ of the aforesaid type
exists, then it can be shown that $\left\langle \hat{B}\right\rangle \le2$.
To see this, consider $\hat{A}_{1}=\hat{a}_{1}\otimes\hat{\mathbb{I}}$
and $\hat{A}_{2}=\hat{\mathbb{I}}\otimes\hat{b}_{1}$, so that if
$\hat{C}:=\hat{A}_{1}\hat{A}_{2}$ then property (2) entails that
$m(\hat{C})=m(\hat{A}_{1})m(\hat{A}_{2})$, viz. $m(\hat{a}_{1}\otimes\hat{b}_{1})=m(\hat{a}_{1}\otimes\hat{\mathbb{I}})m(\hat{\mathbb{I}}\otimes\hat{b}_{1})$.
Note that $m(\hat{*})$, may depend on the state $\left|\psi\right\rangle $
and the hidden variables%
\begin{comment}
, where $\hat{*}$ is taken to represent $\hat{B}_{k}$s which can
be generalised to all $\hat{a}_{i}$s and $\hat{b}_{j}$s
\end{comment}
. Therefore, we average over multiple measurements, which corresponds
to averaging over the hidden variables. This is expressed by $\left\langle m(\hat{a}_{1}\otimes\hat{b}_{1})\right\rangle =\left\langle m(\hat{a}_{1}\otimes\hat{\mathbb{I}})\right\rangle \left\langle m(\hat{\mathbb{I}}\otimes\hat{b}_{1})\right\rangle $.
One can repeat this argument for each term in $\left\langle \hat{B}\right\rangle $.
Using the fact that $-1\le m(\hat{*})\le1$, and that the extrema
will occur only at the extreme values of $m$, one can plugin $\pm1$
for each $\left\langle m(\hat{*})\right\rangle $ to obtain $\left\langle \hat{B}\right\rangle \le2$.
Quantum mechanically, one can show that $\left\langle \hat{B}\right\rangle =2\sqrt{2}$
for the state $\left|\psi\right\rangle =\left(\left|+-\right\rangle -\left|-+\right\rangle \right)/\sqrt{2}$,
where $\hat{\sigma}_{x}\left|\pm\right\rangle =\pm\left|\pm\right\rangle $,
and $\hat{\sigma}_{x,y,z}$ are the Pauli matrices, given in the z-basis
as 
\[
\hat{\sigma}_{x}\doteq\left[\begin{array}{cc}
0 & 1\\
1 & 0
\end{array}\right],\,\hat{\sigma}_{y}\doteq\left[\begin{array}{cc}
0 & -i\\
i & 0
\end{array}\right],\,\hat{\sigma}_{z}\doteq\left[\begin{array}{cc}
1 & 0\\
0 & -1
\end{array}\right].
\]
The observables are $\hat{a}_{1}=\hat{\sigma}_{z}$ and $\hat{a}_{2}=\hat{\sigma}_{x}$,
while $\hat{b}_{1}=-\frac{\hat{\sigma}_{z}+\hat{\sigma}_{x}}{\sqrt{2}}$
and $\hat{b}_{2}=\frac{\hat{\sigma}_{z}-\hat{\sigma}_{x}}{\sqrt{2}}$.
Since quantum mechanics violates the inequality, $\left\langle \hat{B}\right\rangle \le2$,
we conclude that the map $m$ can not simultaneously exist and be
consistent with QM.
\end{proof}
